---
title: "Car Thefts"
---

Follow along with the demo to learn how to calculate standard deviation and how to create a histogram of the data using R and the Tidyverse. Get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
library(dplyr)
```


### Task 2: Load the car thefts dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_jan24.csv


```{r}
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_jan24.csv")
```

### Task 3: Calculate the standard deviation of the number of car thefts, along with the mean

```{r}
car_thefts |> summarize(mean = mean(count), sd = sd(count))
```


### Task 4: Create a histogram of the number of car thefts

```{r}
car_thefts |> 
  ggplot() + 
  geom_histogram(aes(x = count), binwidth = 2) +
  geom_vline(aes(xintercept = mean(count)), color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(count) - sd(count)), color = "blue", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(count) + sd(count)), color = "blue", linetype = "dashed", size = 1)
```



### Task 5: Add a line to the histogram that shows the mean of the number of car thefts, and two other lines to show one standard deviation in both directions


### Task 6: Describe your results

Is the distribution normal or skewed in one direction? Write a sentence in which you describe the range of the data covered by one standard deviation from the mean on both sides - how much of the data falls in that range?

I would say this is a normally distributed data chart. The majority (I'd guess around 60-70%) of the data points fall close to the mean (between the two blue lines) and there isn't a bunch of random spikes in data anywhere. 

When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.




### How likely is a specific car theft month?

Let's pick the month with the highest total and walk through the process: calculate its z-score, use `pnorm()` to find the probability, then check against the actual data.
```{r pnorm-specific-month}
# Find the month with the most car thefts
highest_month <- car_thefts %>%
  filter(count == max(count))

cat(sprintf("Month: %s\n", highest_month$month))
cat(sprintf("Total thefts: %d\n", highest_month$count))

# Step 1: Calculate the z-score
z <- (highest_month$count - mean(car_thefts$count)) / sd(car_thefts$count)
cat(sprintf("\nZ-score: %.2f\n", z))

# Step 2: Use pnorm() to find the probability of a month this high or higher
prob_above <- 1 - pnorm(z)
cat(sprintf("pnorm() estimate - probability of a month this high or higher: %.4f (%.2f%%)\n", prob_above, prob_above * 100))

# Step 3: Check against the actual data
actual_prob <- sum(car_thefts$count >= highest_month$count) / nrow(car_thefts)
cat(sprintf("Actual data - proportion of months this high or higher: %.4f (%.2f%%)\n", actual_prob, actual_prob * 100))
```

The `pnorm()` estimate and the actual proportion won't match perfectly — `pnorm()` assumes a normal distribution, and real data is never perfectly normal. But they should be in the same ballpark, and the comparison helps us see how well the normal model fits.



