---
title: "JOUR405: Statistics for Journalists"
subtitle: "Midterm Exam"
name: "Jack Zwirn"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# JOUR405: Statistics for Journalists
## Midterm Exam

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences.

## Load the tidyverse

```{r}
library(tidyverse)
```

---

## Part 1: Restaurant Health Inspections (25 points)

The first dataset contains restaurant health inspection scores from Montgomery County. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv`

The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible, the compliance score, and the grade.

### Task 1: Read the data (2 points)

```{r}
data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
```

### Task 2: Calculate the mean, standard deviation, minimum, and maximum of compliance scores. Save the results in a new dataframe called `compliance_stats`. (2 points for each)

```{r}
compliance_stats <- data |>
  summarize(mean_compliance = mean(compliance_score) ,
min_compliance = min(compliance_score) ,
max_compliance = max(compliance_score) ,
sd_compliance = sd(compliance_score)
)
```

### Task 3: Create a histogram of the compliance scores with a vertical line showing the mean. Include a descriptive title and label the x and y axes using `labs()`. (10 points)

```{r}
data|> 
  ggplot() + 
  geom_histogram(aes(x = compliance_score), binwidth = 5) +
  geom_vline(aes(xintercept = mean(compliance_score)), color = "red", linetype = "dashed", size = 1) +
  labs(title="Compliance Scores - Restaurant Health Inspections")
```

### Task 4: Write 2–3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What is newsworthy, and what would your next reporting step be? (5 points)


The mean and majority of the compliance scores are high, meaning that the majority of the resturaunts easily passed their health inspections. With a standard deviation of 5.82, that shows that the majority of the resturaunts are clustered around the mean as well. However, my story idea is to cover which and why certain resturaunts faired as poorly as they did. Some were as low as in the 40s, which is very bad for compliance.
---

## Part 2: Daily Transit Ridership — Z-Scores and Sampling (30 points)

Your editor wants you to identify unusual days in WMATA's daily ridership data that might be worth investigating as story leads. The dataset contains daily bus and rail totals. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv`

### Task 5: Load the WMATA ridership data. (2 points)

```{r}
wmata <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")
```

### Task 6: Using `summarize()`, calculate the mean and standard deviation for both bus and rail ridership. Save the results in a dataframe called `wmata_stats`. (5 points)

```{r}
wmata_stats <- wmata |>
  summarize(mean_bus = mean(bus) ,
mean_rail = mean(rail) ,
sd_bus = sd(bus) ,
sd_rail = sd(rail)
)

```

### Task 7: Using `mutate()`, add two new columns to `wmata_daily`: `rail_zscore` (the z-score for rail ridership) and `bus_zscore` (the z-score for bus ridership). Save the result as `wmata_zscores`. (8 points)

```{r}
wmata_zscores <- wmata |>
  mutate(
    rail_zscore = (total - 347987.8) / 100164.3 ,
    bus_zscore = (total - 331293.8) / 88486.88

  )
```

### Task 8: Filter `wmata_zscores` to show only days where the absolute value of `rail_zscore` is greater than 2 (unusually high or low rail ridership). Arrange by the absolute value of the z-score, with the most extreme first. How many unusual days are there? (8 points)

```{r}
wmata_zscores |>
  filter(abs(rail_zscore) > 2) |>
  arrange(desc(abs(rail_zscore)))
```

There are 301 unusual days where the zscore is greater than 2.

### Task 9: Using `set.seed()`, take a random sample of 91 days from `wmata_daily` (roughly 25% of the data). Calculate the mean and standard deviation for bus and rail ridership in the sample. How do the sample statistics compare to the population statistics you calculated in Task 6? Write 1–2 sentences comparing the values and explaining what this suggests about random sampling. (4 points)

```{r}
set.seed(91)

sample_91d <- wmata |> sample_n(91)

sample_91s <- sample_91d |>
summarize(mean_bus = mean(bus) ,
mean_rail = mean(rail) ,
sd_bus = sd(bus) ,
sd_rail = sd(rail)
)

```

While they are not exactly the same, they are close to each another, suggesting that the sample of 25% of the data is sufficient when randomly selected. It would be better to sample a larger portion, however, as the larger the random sample, the closer to being 100% accurate the data will be.

### Task 10: Write 2–3 sentences: What do the unusual days in your z-score analysis suggest as story leads? (3 points)

Why are there so many/few people using the WMATA system on those specific days? The stories could be about if there is any pattern between the outlier dates and what happened on those days.
---

## Part 3: Maryland Car Theft Rates (25 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to find out which counties have the highest rates. The dataset contains car theft counts for Maryland counties in 2023 along with population figures. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv`

### Task 11: Load the car thefts data. (2 points)

```{r}
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
```

### Task 12: Using `mutate()`, add a column called `theft_rate` that calculates the car theft rate per 10,000 residents for each county. Save the result back to `md_car_thefts`. In 1–2 sentences, explain why a per-capita rate is more useful than raw counts for this comparison. (8 points)

```{r}
car_thefts <- car_thefts |>
  mutate(
    theft_rate = thefts_2023/population * 10000
  )
```

Because the population of a county is huge, it is easier to use a per-capita rate because it shows how many thefts per "x" number of people, rather than a raw count, which says "there are x thefts". It makes more sense to readers if the data is more approximated, rather than just a large number.

### Task 13: Calculate the median car theft rate and the total number of car thefts statewide. Show the counties that have rates above the median. (7 points)

```{r}
car_thefts |>
  summarize(
    median_thefts = median(theft_rate) ,
    total_thefts = sum(thefts_2023)
  )
```
Baltimore City, Prince George's County, St. Mary's County, Baltimore County, Dorchester County, Montgomery County, Howard County, Anne Arundel County, Washington County, Charles County, Cecil County, Wicomico County all have rates above the median


### Task 14: Write 2–3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What is the lede of a story based on your findings? (8 points)

The distribution of car thefts across Maryland is fairly normal across most of the state, with some being high and some being low. However, Baltimore City and PG County have rates way higher than the rest. 

Lede: Car thefts in Baltimore City and PG County are high when compared to the rest of the state.

## Part 4: Data Analysis Scenario (20 points)

A city parks department releases data showing that its 12 community swimming pools average 187 visitors per day during the summer season, with a standard deviation of 42 visitors. Your editor suspects that a handful of pools are dramatically underperforming and wants a story. You also learn that the department's figures are based on a random sample of 60 days per pool rather than a complete count.

Write 3–4 sentences for each of the following questions:

### Question A: Given that the mean is 187 and standard deviation is 42, which statistical measures would you prioritize using and why? (7 points)
I would use the mean of 187 visitors over the standard deviation because the average visitor count is something that readers can resonate with easily. The mean show readers how busy the pool is going to be on an average day. The standard deviation shows important information as well. It shows how often the pools are crowded, but for this story, I think the average would be more helpful to most readers.

### Question B: How would you use z-scores to identify which pools are genuinely underperforming, and why is that more useful than simply ranking pools by raw attendance? (7 points)

I would use zscores to find outliers of which pools are performing the worst by taking the total visitors, subtracting the mean, and then dividing the result by the standartd deviation. This is more helpful than ranking pools by attendance because zscores show how far off specific pools are. You can look at the raw numbers, but the zscore shows us in an easier way which pools are way off in comparison to the average. If a pool is over 2 standard deviations away from the mean, then the pool is either performing really well, or really poorly.

### Question C: The department collected its attendance figures using a random sample of 60 days per pool rather than tracking every day. What questions would you ask about the sample before relying on these figures for a story? What concerns would you raise in your story? Use statistical terminology where appropriate. (6 points)

I would look at how much of a normal distribution there is in the random sample. If there is a giant skew to one side, than I would ask for a different sample because they could be trying to skew the data in their favor, despite them saying it was random. I would also ask the department when the data was from and look at the dates in the sample to see if they used days spread out from the year or clustered together in the summer months when pools are being used more frequently. I would also take a smaller sample from their sample to see how close the mean and standard deviations are with half of their data to see if the data remains consistent. I would raise the concern of why are they only taking data on 60 days, instead of every day? Why wouldn't they want the most accurate data possible that we could then take a sample from?

### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your choices

Good luck!
