---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "Jack Zwirn"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

In the class activity, you tested a school board member's claim using data from Montgomery County, Pa. — and found a model that didn't hold up. Now you'll run the same analysis using data from Montgomery County, Md. Your job is to evaluate whether the Maryland data tells a different story, and what that means for how you'd report on the claim.

### Loading the Data

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```

### Task 1: Examine the Data First (2 points)

Before you plot anything, look at the data — you know why by now.

```{r}
summary(schools$`FARMS Pct`)
schools |> arrange(desc(`FARMS Pct`)) |> head(10)
```

Are there any data quality issues like the ones you found in the Pennsylvania data? How does the distribution of FARMS percentages compare between the two counties? Note anything that concerns you or gives you confidence in the data.

There are not data quality issues here as there aren't any schools with the 100% FARMS PCT. The distribution is more spread out in comparison to the data for Montgomery County PA. What makes me like this more in comparison to the PA data is the fact that there is a more even spread of FARMS percentages across the entire data spectrum.


### Task 2: Create a Scatterplot and Compare (5 points)

Replace the REPLACE_ME value with the correct variable name to create a scatterplot of FARMS percentage vs. math scores for Montgomery County, Md.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage, Montgomery County MD")
```

Describe what you see. Then compare this scatterplot specifically to the Pennsylvania one: what differences do you notice in the direction, strength, and spread of the relationship? The two counties share a name and are both suburban — what might explain why they look so different?

There is a negative correlation between an increase in FARMS score pct and the scores on math exams. While not perfect, the general downward trend is clearly shown. Perhaps Montgomery County MD is a generally richer area where the FARMS program is needed less so the schools that do need it perform poorly anyways because their students are more distracted by a family situation or something like that where their focus isn't on their school work. Perhaps the schools with the low FARMS pct have higher math scores because they generally have more food security.


### Task 3: Linear Regression Model (5 points)

Replace the REPLACE_ME value with the correct variable name to run a linear regression model predicting math scores from FARMS percentage.

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)
summary(model)
```

Report the three key statistics: the p-value, the Adjusted R-squared, and the coefficient for FARMS Pct. Then interpret them together — not separately. What does the combination of these three numbers tell you about the relationship between poverty and math scores in Montgomery County, Md.? Is this model worth using? How does it compare to the Pennsylvania model?

p-value: < 0.00000000000000022
adjusted r-squared: 0.6734
coefficient: -0.59511
These numbers tell me that this data is statistically significant and there is a correlation between the two variables. However, this significance is extremely negative meaning that we can strongly reject the null hypothesis. This model is worth using, because it tells us that there is the negative correlation and that it can tell us that these numbers don't have any real effects on one another. This model is much more significant compared to the PA model because this one gives us a difinitive answer.


### Task 4: Identifying Schools with the Biggest Residuals (6 points)

Replace the REPLACE_ME values with the correct variable names to calculate predicted values, residuals, and absolute residuals for each school. Then arrange the schools in descending order of absolute residuals.

```{r}
schools <- schools |>
  mutate(
    predicted = predict(model),
    residual = `% Met or Exceeded Math Expectations` - predicted,
    abs_residual = abs(residual)
  ) |>
  arrange(desc(abs_residual))

head(schools |> select(1, `FARMS Pct`, `% Met or Exceeded Math Expectations`, predicted, residual), 10)
```

Identify the school with the largest positive residual (doing much better than the model predicts) and the school with the largest negative residual (doing much worse). For each school, find its zip code and look it up on [CensusReporter.org](https://censusreporter.org).

For each of the two schools, answer:
- What does the demographic and economic profile of the zip code look like?
- Does anything in that profile help explain why the school is over- or under-performing the model?
- What would your first reporting question be to the principal of that school?

Largest positive residual: Fox Chapel Elementary - 26,406 - lower avg income, 9.8% under poverty line
Largest negative residual: Germantown Elementary - 62,738 - higher avg income, 8.3% under poverty line
Both are pretty much the same when it comes to their demographics. 50/50 male female split, avg age of abt 37, 23% of residents under 18 years old
Nothing pops out at me for why the schools would be over or underperforming outside of the sheer number of people in each area. For Germantown Elementary, the resources have to be spread out more across more total people while at Fox Chapel there is less total people meaning that the resources can be given to each student more individually.
My first question to the Germantown principal would be "How do you plan to better divy your resources to help each student improve overall?"
My first question to the Fox Chapel principal would be "What do you feel that you are doing well to help your students have higher grades?"


### Task 5: Evaluate the Model and Write the Pitch (7 points)

You've now run this analysis on two counties. Return to the school board member's claim: that elementary schools with fewer students on free and reduced-price meals do better on standardized math tests.

Write a short story pitch of 4-6 sentences that addresses all of the following: Does the Maryland data support or contradict the claim? How does it compare to what you found in Pennsylvania? What would you ask the school board member if you called them for comment? And what's the most interesting specific finding from your residuals analysis: which school would you want to visit first, and why?

The Maryland data supports the claim from the school board member. It is in stark contrast to the PA data. I would ask the school board rep about why they think there is such a difference between the two counties, considering the data we looked at was the same. The most interesitgn finding to me is that Stedwick Elementary had an 8.3% met or exceeded expectations mark when it was predicted to be in the 30% range. Why is the result SO low there?



