---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

Most of the categories fall into categories themselves. Assault, B&E, robbery and theft. There are a few exceptions such as "accident" or "homicide". 
For reporters, all of these categories are important. However, the ones that are more important are the categories that have frequent or higher numbers compared to their "expected" frequencies. For this set of data, one story could be about how "B&E Commercial" crimes are more common in December, around the holidays. Another could be about the low percentage of "Auto Stolen & Recovered" compared to the number of "Auto Stolen". 
I don't think any of these categories make the data harder to summarize. All of these categories are valuable, and you can't really combine any of them to make it "easier".

### Task 4: Calculating Averages

Calculate the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.


```{r}
pg_crime_averages <- pg_crime |> mutate(total_incidents = Jan + Feb + Mar + Apr + May + Jun + Jul + Aug + Sep + Oct + Nov + Dec)
```

```{r}
pg_crime_averages2 <- pg_crime_averages |> mutate(average_incidents = total_incidents/12)
```

### Task 5: Describe the Averages

Examine the new data-frame, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

Two of the most common crimes were "Accident" and "Accident with Impound". These are ones that I figured would be among the most common, as car wrecks are unfortunately a common occurrence.

Homicides were one of the least common crimes. I take that to be a good thing, but averaging just under six homicides per month is still too high. This is where you need to take into account what the crime is. Certain crimes, like homicides, should be a less frequent occurrence. However, their "high" numbers are seemingly low in comparison.

This data tells me that there is a large number of crimes committed per year in PG County and they are seemingly all at "normal" rates. I say normal because none of the categories jump out at me as being egregiously high or low.


### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime_averages2 |> mutate(june_change = (Jun - average_incidents) / average_incidents *100, july_change = (Jul - average_incidents) / average_incidents *100, august_change = (Aug - average_incidents) / average_incidents *100)
```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

The average number of crimes committed stayed fairly the same in the summer months. While some categories saw a large change in one month, the other two summer months showed the opposite or small changes. The pattern is that crimes will always fluctuate throughout the year. Sometimes crime numbers go up, whereas other times crime numbers will decrease. 

### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Auto thefts more commonly result in no outcome, rather than a satisfying conclusion for the car owner. 

Of the roughly 5,000 auto thefts last year, only 10% were returned to their rightful owners. Approximately 400 cars were stolen each month, and an average of 40 were returned. Despite their best efforts, police fail to return the stolen property more often than not, causing victims to lose their belongings.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
