---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2025.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

Some employees salaries are much higher or lower than you would expect. For example, Luis Hinojos, a senior policy advisor makes $83,500 per year, while Jacalynne Klopp, an "adivsor", makes $225,700 per year.
There are also a few employees who are shown to make $0 annually. This will skew the analysis and could create some inaccuracies. 

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |>
  summarize(
    mean_salary = mean(SALARY),
    median_salary = median(SALARY)
  )
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |>
  summarize(
    mean_salary = mean(SALARY),
    median_salary = median(SALARY),
    min_salary = min(SALARY),
    max_salary = max(SALARY)
  )
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The mean and median salaries show me that most White House employees earn over $100,000 annually. However, that data is skewed by some outliers. A few employees are shown to earn $0 per year, and some others earn up to $225,700 per year. These outliers show us that while helpful, the mean and median aren't 100% reliable information without going through and taking out the outliers in the data set.

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

My best story idea is about why certain White House employees earn $0 annually. The data didn't list them as interns or in an introductory role. At least one of them was listed as a "senior advisor". Other "advisors" and "senior advisors" get paid so why don't all of them? One "senior advisor" gets paid just under $200,000 per year. 
My next step for developing this story is to reach out to the White House press office and ask for clarification. Was this an error? Are there actually White House employees who don't get paid?


