---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv") |> clean_names()
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |> mutate(violent_rate_2019 = violent2019/pop2019 * 10000, property_rate_2019 = property2019/pop2019 *10000, violent_rate_2020 = violent2020/pop2020 *10000, property_rate_2020 = property2020/pop2020 *10000)
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

The higher the population, the higher the crime rates are. For example, Baltimore and Annapolis have higher populations and crime rates compared to smaller towns such as Belair and Berlin. Cities tend to have higher populations, leading to more crime. The story is: the more populated an area, the more likely it is for crime rates to be high.

### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities |> mutate(violent_pct_change = (violent2020- violent2019)/ violent2019 * 100, property_pct_change = (property2020 - property2019)/property2019 *100)
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

The majority of towns and cities in Maryland have negative percent change results, meaning that crime is decreasing. However, what is more interesting is the outliers. Boonsboro and Crisfield are two examples of this. Both derive from the norm of negative percent change, instead increasing by a large margin. One story idea could be delving into why some of these smaller towns like Boonsboro and Crisfield have such high crime rates. On the flip side, some towns with very small populations have zero reported crimes at all.

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Small Maryland towns reported zero violent or property crimes in 2019 and 2020.

Some smaller towns in Maryland have seen a complete absence of crime in recent years. Two examples of this are the towns of Barclay and Barnesville. Both towns reported a nonexistence of crime in both 2019 and 2020. The combined population of 347 people between the two towns plays a huge role in that result.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.


### Task 10

Using the 2020 violent crime rate, calculate the percentiles using `ntile()`. Save the results to a new dataframe called `md_cities_percentiles`. How many cities are above the 90th percentile?

Calculate the Interquartile Range of the 2020 violent crime rate using `summarize()` and R's `IQR()` function. The syntax is similar to how `mean()` works.

Do the results change your best story idea from the previous assignment? Why or why not?

```{r}
md_cities_percentiles <- md_cities |> mutate(percentile = ntile(violent2020, 100))
```

11 cities are above the 90th percentile

```{r}
md_cities_percentiles |> 
  summarize(
    IQR_violent2020 = IQR(violent2020)
  )
```

The results does change my story idea. While Barclay and Barnesville are both in the 1st percentile, it is still an interesting story that their small population has helped them have zero crime in their towns. However, there are definitely some new ideas that can be found from this new data such as Ocean City having such a low population but being in the 94th percentile. This would be a very interesting story because their small population of just under 7,000 resulting in high crime rates. Maybe because it is a popular tourist destination, that results in some of the higher numbers.



