---
title: "final_project"
output: html_document
---
The dataframe contains data from 1954-2013. The data shows the different sports that were featured on the cover of Sports Illustrated over that 59 year span. The columns that are specific to my project are: baseball, basketball, boxing, american football and ice hockey. This project will look at their coverage on the magazine cover over time, being the "Big Four" sports in the US. How do they fluctuate in comparison to each other? How does this reflect the popularity of each sport? I will exclude the data from 1954 because they only printed 20 issues that year, rather than the 52 or 53 from every other year in the dataset.



Hypothesis testing to see the correlation over time of how commonly they were used. Hypotheses: Football became more common of a cover sport over time. Baseball and basketball remained relatively flat but on a slight decline. Ice hockey was kind of randomly thrown in there over time when deemed right. To help with this, I will create bar charts to look at whether or not these sports have a normal distribution over time, if they are skewed towards the earlier or later years, or if it is random.

Fox started broadcasting sports in 1994. How did the broadcasting company bringing more eyes to live sports affect who was on the cover of the magazine? I also want to do the same thing, but for ESPN. They started in 1979.

I want to do a similar comparison, but using the year 2000. I chose this year as well because I want to see if the "digital age" had any affect on which sports were featured more frequently.

Linear regression would be the perfect way to test the above comparisons (FOX, ESPN, 2000). I want to maybe do a similar comparison to what we did in class with the FARMS data to see if there is a correlation over time.

Outlier analysis. Were there any years where there was little to no coverage of any of these sports? Or on the flip side, were there any years where there was a much larger number of issues made with the sport featured? What could have caused these outliers?

I also want to turn each data entry into a percentage of the total issues for that year. How do they compare to each other?




After starting the process of working through the project, I am starting to figure out what I truly want to do with this project. I already have the percentages per year done for each sport as well as the overall percentage totals. These percentages helped me determine which sports are the most commonly covered. I wonder how I can take this a step further, but I am unsure how to do so yet. 

I started to create bar charts for each sport, and I want to add lines to show where FOX/ESPN/2000 are and then see those comparisons. These bar charts will also show me the general trend over time for each of these sports and how frequently they were represented on Sports Illustrated covers over time. 

The final step will be to make linear regression charts to show the pct share over time. I know the general trend will be found by the bar charts, but I will use the linear regression charts to determine whether or not the result is significant by finding the p-value. Then finding the z-score to determine whether or not there are any outliers.





Start of project:


```{r}

library(tidyverse)
library(janitor)
library(dplyr)


covers <- read_csv("SI-Cover-by-Sport-DFE.csv")

covers <- covers %>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))

```

```{r}
covers <- read_csv("SI-Cover-by-Sport-DFE.csv") |> clean_names()  |> rename (year = unnamed_0) # becomes: american_football, ice_hockey, grand_total, etc.

covers2 <- covers |>
  filter(year >= 1955)
covers2 <- covers %>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))

```

```{r}
covers2 <- covers2 |>
   mutate(across(c(american_football, baseball, boxing, basketball, ice_hockey),
                ~replace_na(., 0))) |>
  pivot_longer(c(american_football, baseball, boxing, basketball, ice_hockey),
               names_to = "sport", values_to = "count") |>
  mutate(share = count / grand_total)
  
```

```{r}
covers2 <- covers2 |>
  filter(year >= 1955)
```

```{r}
covers2 <- covers2 |>
  select(year, grand_total, sport, count, share)
```

The above code is to filter the data to be relevant for my project. I created a second copy of the chart so I still have the original dataframe. "Covers2" only shows the data form the 5 sports that I am looking at for this project: american football, basketball, baseball, ice hockey and boxing. I also filtered the data to only show 1955 and after to avoid the 1954 anomaly of half the number of magazine issues.

I filtered it by "share" to see what the highest shares of covers by sport were from 1955-2013. The highest percentage is 46% for American Football in 2005. In fact, each of the top 12 cover share percentages are all American Football, showing it's prominence and dominance in the sports landscape. The next 12 are all tied at 17 covers in the year. 8 of them are American Football. This means that of the top 24, 16.66% of them were not American Football, meaning 83.33% of them were. This was 4 of the 24. 3 of them were basketball and 1 was baseball.

It is the same story when looking at the overall numbers. 27% of all Sports Illustrated covers from 1955-2013 had American Football represented on the cover. 19% baseball. 18% basketball. 4% boxing. 3% ice hockey. 71% of all Sports Illustrated covers were one of these 5 sports, which are arguably the top 5 sports in US history. Other sports like NASCAR, golf and soccer have their popularity too, but when considering all of US history, the 5 sports I mentioned above are at the top of the charts.


```{r}
basketball <- covers2 %>%
  filter(sport == "basketball")
baseball <- covers2 %>%
  filter(sport == "baseball")
american_football <- covers2 %>%
  filter(sport == "american_football")
ice_hockey <- covers2 %>%
  filter(sport == "ice_hockey")
boxing <- covers2 %>%
  filter(sport == "boxing")
```

Above: This is me creating separate dataframes, categorized by each sport so I can get a better understanding of how each of my tests will affect each sport individually. I will create my charts individually by sport so I can easily see the differences.

```{r, fig.width = 12, fig.height = 5}
vlines <- data.frame(
  year = c("1979", "1994", "2000"),
  event = c("Start of ESPN", "Start of FOX", "Start of Digital Age")
)

ggplot(boxing[boxing$year != "Grand Total", ], aes(x = year, y = count, group = 1)) +
  geom_line() +
  geom_vline(data = vlines, aes(xintercept = year, color = event), linetype = "solid") +
  scale_color_manual(
    name = "Key Events",
    values = c("Start of ESPN" = "red", "Start of FOX" = "blue", "Start of Digital Age" = "green")
  ) +
  labs(
    x = "Year",
    y = "Number of Boxing Covers",
    title = "Sports Illustrated Covers: Boxing Coverage Over Time"
  ) +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1, size = 9)
  )
```
Looking at this line chart, I see a steep downward trend of coverage, especially since Muhammad Ali's career ended in 1981. This correlates with the start of ESPN as they almost immediately started covering boxing when they started up, starting their boxing coverage in 1980. ESPN never aired a fight with Ali in it. After Ali retired, the popularity of the sport decreased, causing a lower amount of coverage on Sports Illustrated covers. Ali holds the second-most appearances on the cover, behind only Michael Jordan, being featured on the cover 40 times between 1963-2016. Over the course of Ali's career, there were 60 boxing features on the Sports Illustrated cover. One can make the reasonable assumption that Ali was the cover athlete for the majority of those 60 throughout his career. Being the face of the sport, one can guess that once he retired, the coverage went down as a result. The comparisons for ESPN and FOX aren't very helpful for boxing as FOX never broadcasted boxing and ESPN got into the sport as it was at the end of its peak, but they still split up the data to show that the sport has declined in popularity. For the third line, 2000, this shows the severity of the decline as there is a total of 3 covers featuring boxing after the year 2000.

```{r}
boxing_filtered <- boxing[boxing$year != "Grand Total", ]
boxing_filtered$year <- as.numeric(boxing_filtered$year)
boxing_filtered$count <- as.numeric(boxing_filtered$share)

# Before/After ESPN (1979)
boxing_filtered$espn <- ifelse(boxing_filtered$year < 1979, "Before ESPN", "After ESPN")
aggregate(count ~ espn, data = boxing_filtered, FUN = mean)

# Before/After FOX (1994)
boxing_filtered$fox <- ifelse(boxing_filtered$year < 1994, "Before FOX", "After FOX")
aggregate(count ~ fox, data = boxing_filtered, FUN = mean)

# Before/After Digital Age (2000)
boxing_filtered$digital <- ifelse(boxing_filtered$year < 2000, "Before Digital", "After Digital")
aggregate(count ~ digital, data = boxing_filtered, FUN = mean)
```
These comparisons are here to show the share percentages per year for before/after each of the comparison lines. There isn't much to take away from these comparisons as the lines aren't very helpful for this particular sport. The digital age one, however, reflects what I mentioned earlier about the severe lack of coverage post 2000. 

```{r}
model_filtered <- lm(`year` ~ `share`, data = boxing)
summary(model_filtered)
```
The linear regression model shows me that this comparison is statistically significant. With an extremely low p-value, this tells me that there is a relationship between the year and the percentage share of covers using boxing. The extremely low coefficient tells me that there is a STEEP decline in boxing coverage over time.


