---
title: "final_project"
output: html_document
---
The dataframe contains data from 1954-2013. The data shows the different sports that were featured on the cover of Sports Illustrated over that 59 year span. The columns that are specific to my project are: baseball, basketball, boxing, american football and ice hockey. This project will look at their coverage on the magazine cover over time, being the "Big Four" sports in the US. How do they fluctuate in comparison to each other? How does this reflect the popularity of each sport? I will exclude the data from 1954 because they only printed 20 issues that year, rather than the 52 or 53 from every other year in the dataset.



Hypothesis testing to see the correlation over time of how commonly they were used. Hypotheses: Football became more common of a cover sport over time. Baseball and basketball remained relatively flat but on a slight decline. Ice hockey was kind of randomly thrown in there over time when deemed right. To help with this, I will create bar charts to look at whether or not these sports have a normal distribution over time, if they are skewed towards the earlier or later years, or if it is random.

Fox started broadcasting sports in 1994. How did the broadcasting company bringing more eyes to live sports affect who was on the cover of the magazine? I also want to do the same thing, but for ESPN. They started in 1979.

I want to do a similar comparison, but using the year 2000. I chose this year as well because I want to see if the "digital age" had any affect on which sports were featured more frequently.

Linear regression would be the perfect way to test the above comparisons (FOX, ESPN, 2000). I want to maybe do a similar comparison to what we did in class with the FARMS data to see if there is a correlation over time.

Outlier analysis. Were there any years where there was little to no coverage of any of these sports? Or on the flip side, were there any years where there was a much larger number of issues made with the sport featured? What could have caused these outliers?

I also want to turn each data entry into a percentage of the total issues for that year. How do they compare to each other?




After starting the process of working through the project, I am starting to figure out what I truly want to do with this project. I already have the percentages per year done for each sport as well as the overall percentage totals. These percentages helped me determine which sports are the most commonly covered. I wonder how I can take this a step further, but I am unsure how to do so yet. 

I started to create bar charts for each sport, and I want to add lines to show where FOX/ESPN/2000 are and then see those comparisons. These bar charts will also show me the general trend over time for each of these sports and how frequently they were represented on Sports Illustrated covers over time. 

The final step will be to make linear regression charts to show the pct share over time. I know the general trend will be found by the bar charts, but I will use the linear regression charts to determine whether or not the result is significant by finding the p-value. Then finding the z-score to determine whether or not there are any outliers.





Start of project:


```{r}

library(tidyverse)
library(janitor)
library(dplyr)

```

```{r}
covers <- read_csv("SI-Cover-by-Sport-DFE.csv") |> clean_names()  |> rename (year = unnamed_0) # becomes: american_football, ice_hockey, grand_total, etc.

covers2 <- covers |>
  filter(year >= 1955) |> 
  mutate(across(where(is.numeric), ~replace_na(., 0))) |> 
  mutate(year = as.numeric(year))

```

```{r}
covers2 <- covers2 |>
   mutate(across(c(american_football, baseball, boxing, basketball, ice_hockey),
                ~replace_na(., 0))) |>
  pivot_longer(c(american_football, baseball, boxing, basketball, ice_hockey),
               names_to = "sport", values_to = "count") |>
  mutate(share = count / grand_total)
  
```

```{r}
covers2 <- covers2 |>
  select(year, grand_total, sport, count, share)
```

The above code is to filter the data to be relevant for my project. I created a second copy of the chart so I still have the original dataframe. "Covers2" only shows the data form the 5 sports that I am looking at for this project: american football, basketball, baseball, ice hockey and boxing. I also filtered the data to only show 1955 and after to avoid the 1954 anomaly of half the number of magazine issues.

I filtered it by "share" to see what the highest shares of covers by sport were from 1955-2013. The highest percentage is 46% for American Football in 2005. In fact, each of the top 12 cover share percentages are all American Football, showing it's prominence and dominance in the sports landscape. The next 12 are all tied at 17 covers in the year. 8 of them are American Football. This means that of the top 24, 16.66% of them were not American Football, meaning 83.33% of them were. This was 4 of the 24. 3 of them were basketball and 1 was baseball.

It is the same story when looking at the overall numbers. 27% of all Sports Illustrated covers from 1955-2013 had American Football represented on the cover. 19% baseball. 18% basketball. 4% boxing. 3% ice hockey. 71% of all Sports Illustrated covers were one of these 5 sports, which are arguably the top 5 sports in US history. Other sports like NASCAR, golf and soccer have their popularity too, but when considering all of US history, the 5 sports I mentioned above are at the top of the charts.


```{r}
basketball <- covers2 %>%
  filter(sport == "basketball")
baseball <- covers2 %>%
  filter(sport == "baseball")
american_football <- covers2 %>%
  filter(sport == "american_football")
ice_hockey <- covers2 %>%
  filter(sport == "ice_hockey")
boxing <- covers2 %>%
  filter(sport == "boxing")
```

Above: This is me creating separate dataframes, categorized by each sport so I can get a better understanding of how each of my tests will affect each sport individually. I will create my charts individually by sport so I can easily see the differences.


BOXING:

```{r, fig.width = 12, fig.height = 5}
vlines <- data.frame(
  year = c("1979", "1994", "2000"),
  event = c("Start of ESPN", "Start of FOX", "Start of Digital Age")
)

ggplot(boxing[boxing$year != "Grand Total", ], 
       aes(x = as.numeric(year), y = share, group = 1)) +
  geom_line() +
  geom_vline(data = vlines, aes(xintercept = as.numeric(year), color = event), linetype = "solid") +
  scale_color_manual(
    name = "Key Events",
    values = c("Start of ESPN" = "red", "Start of FOX" = "blue", "Start of Digital Age" = "green")
  ) +
  scale_x_continuous(breaks = seq(1955, 2013, by = 5)) +
  labs(
    x = "Year",
    y = "Number of Boxing Covers",
    title = "Sports Illustrated Covers: Boxing Coverage Over Time"
  ) +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1, size = 9)
  )
```
Looking at this line chart, I see a steep downward trend of coverage, especially since Muhammad Ali's career ended in 1981. This correlates with the start of ESPN as they almost immediately started covering boxing when they started up, starting their boxing coverage in 1980. ESPN never aired a fight with Ali in it. After Ali retired, the popularity of the sport decreased, causing a lower amount of coverage on Sports Illustrated covers. Ali holds the second-most appearances on the cover, behind only Michael Jordan, being featured on the cover 40 times between 1963-2016. Over the course of Ali's career, there were 60 boxing features on the Sports Illustrated cover. One can make the reasonable assumption that Ali was the cover athlete for the majority of those 60 throughout his career. Being the face of the sport, one can guess that once he retired, the coverage went down as a result. The comparisons for ESPN and FOX aren't very helpful for boxing as FOX never broadcasted boxing and ESPN got into the sport as it was at the end of its peak, but they still split up the data to show that the sport has declined in popularity. For the third line, 2000, this shows the severity of the decline as there is a total of 3 covers featuring boxing after the year 2000.

```{r}
boxing_filtered <- boxing[boxing$year != "Grand Total", ]
boxing_filtered$year <- as.numeric(boxing_filtered$year)
boxing_filtered$count <- as.numeric(boxing_filtered$share)

# Before/After ESPN (1979)
boxing_filtered$espn <- ifelse(boxing_filtered$year < 1979, "Before ESPN", "After ESPN")
aggregate(count ~ espn, data = boxing_filtered, FUN = mean)

# Before/After FOX (1994)
boxing_filtered$fox <- ifelse(boxing_filtered$year < 1994, "Before FOX", "After FOX")
aggregate(count ~ fox, data = boxing_filtered, FUN = mean)

# Before/After Digital Age (2000)
boxing_filtered$digital <- ifelse(boxing_filtered$year < 2000, "Before Digital", "After Digital")
aggregate(count ~ digital, data = boxing_filtered, FUN = mean)
```
These comparisons are here to show the share percentages per year for before/after each of the comparison lines. There isn't much to take away from these comparisons as the lines aren't very helpful for this particular sport. The digital age one, however, reflects what I mentioned earlier about the severe lack of coverage post 2000. 

```{r}
model_filtered <- lm(`year` ~ `share`, data = boxing)
summary(model_filtered)
```
The linear regression model shows me that this comparison is statistically significant. With an extremely low p-value, this tells me that there is a relationship between the year and the percentage share of covers using boxing. The extremely low coefficient tells me that there is a STEEP decline in boxing coverage over time.





ICE HOCKEY:

```{r, fig.width = 12, fig.height = 5}
vlines <- data.frame(
  year = c("1979", "1994", "2000"),
  event = c("Start of ESPN", "Start of FOX", "Start of Digital Age")
)

ggplot(ice_hockey[ice_hockey$year != "Grand Total", ], 
       aes(x = as.numeric(year), y = share, group = 1)) +
  geom_line() +
  geom_vline(data = vlines, aes(xintercept = as.numeric(year), color = event), linetype = "solid") +
  scale_color_manual(
    name = "Key Events",
    values = c("Start of ESPN" = "red", "Start of FOX" = "blue", "Start of Digital Age" = "green")
  ) +
  scale_x_continuous(breaks = seq(1955, 2013, by = 5)) +
  labs(
    x = "Year",
    y = "Number of Ice Hockey Covers",
    title = "Sports Illustrated Covers: Ice Hockey Coverage Over Time"
  ) +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1, size = 9)
  )
```
Looking at this data, I see a sport that never got much coverage in comparison to the total number of issues in any given year, but remained relatively consistent at giving a few each year. You can see a giant uptick in 1980, which is when the "Miracle on Ice" team won gold for the US in the Winter Olympics. This warranted an increase in coverage on the cover of Sports Illustrated. Aside from that spike, there was a few year span in the early 2000s where there was zero coverage. This can be choked up to the 2004/2005 lockout, with no action on the ice for a long period of time. Hockey has always seemingly been in the 4 spot behind the NFL, MLB and NBA within the US and this data reflects that.

```{r}
ice_hockey_filtered <- ice_hockey[ice_hockey$year != "Grand Total", ]
ice_hockey_filtered$year <- as.numeric(ice_hockey_filtered$year)
ice_hockey_filtered$count <- as.numeric(ice_hockey_filtered$share)

# Before/After ESPN (1979)
ice_hockey_filtered$espn <- ifelse(ice_hockey_filtered$year < 1979, "Before ESPN", "After ESPN")
aggregate(count ~ espn, data = ice_hockey_filtered, FUN = mean)

# Before/After FOX (1994)
ice_hockey_filtered$fox <- ifelse(ice_hockey_filtered$year < 1994, "Before FOX", "After FOX")
aggregate(count ~ fox, data = ice_hockey_filtered, FUN = mean)

# Before/After Digital Age (2000)
ice_hockey_filtered$digital <- ifelse(ice_hockey_filtered$year < 2000, "Before Digital", "After Digital")
aggregate(count ~ digital, data = ice_hockey_filtered, FUN = mean)
```
After ESPN started, the share of covers featuring hockey on the cover dropped by a significant margin. Aside from the uptick in 1980 with the Miracle on Ice, overall, coverage went down. ESPN began their NHL coverage right as they started as a company. There is a similar pattern for both the FOX and digital age lines as well with the share of coverage going down compared to before they started. FOX was only a broadcast partner for the NHL from 1995 to 1999 and in those years the share stayed low before it made its final dip after the digital age. 

```{r}
model_filtered2 <- lm(`year` ~ `share`, data = ice_hockey)
summary(model_filtered2)
```

This linear regression model shows that there is a significant decrease in NHL coverage over time. There is a larger decrease in coverage over time than compared to the boxing coverage regression over time. Both of these sports had their time at the top, but was short lived and dropped off sharply after the fact.



BASKETBALL:

```{r, fig.width = 12, fig.height = 5}
vlines <- data.frame(
  year = c("1979", "1994", "2000"),
  event = c("Start of ESPN", "Start of FOX", "Start of Digital Age")
)

ggplot(basketball[basketball$year != "Grand Total", ], 
       aes(x = as.numeric(year), y = share, group = 1)) +
  geom_line() +
  geom_vline(data = vlines, aes(xintercept = as.numeric(year), color = event), linetype = "solid") +
  scale_color_manual(
    name = "Key Events",
    values = c("Start of ESPN" = "red", "Start of FOX" = "blue", "Start of Digital Age" = "green")
  ) +
  scale_x_continuous(breaks = seq(1955, 2013, by = 5)) +
  labs(
    x = "Year",
    y = "Number of Basketball Covers",
    title = "Sports Illustrated Covers: Basketball Coverage Over Time"
  ) +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1, size = 9)
  )
```
This chart shows an increase over time in basketball coverage. While there is a drop off at the end, the overall trend is up. The increase in the late 1980s/early 1990s corresponds pretty closely with Michael Jordan's career taking off. With Jordan being one of the top NBA players of all time (and he is #1 in my mind) he has been featured on the cover 50 times from 1983 to 2013, a record for any individual athlete.

```{r}
basketball_filtered <- basketball[basketball$year != "Grand Total", ]
basketball_filtered$year <- as.numeric(basketball_filtered$year)
basketball_filtered$count <- as.numeric(basketball_filtered$share)

# Before/After ESPN (1979)
basketball_filtered$espn <- ifelse(basketball_filtered$year < 1979, "Before ESPN", "After ESPN")
aggregate(count ~ espn, data = basketball_filtered, FUN = mean)

# Before/After FOX (1994)
basketball_filtered$fox <- ifelse(basketball_filtered$year < 1994, "Before FOX", "After FOX")
aggregate(count ~ fox, data = basketball_filtered, FUN = mean)

# Before/After Digital Age (2000)
basketball_filtered$digital <- ifelse(basketball_filtered$year < 2000, "Before Digital", "After Digital")
aggregate(count ~ digital, data = basketball_filtered, FUN = mean)
```
ESPN started broadcasting the NBA in 1983, corresponding with Jordan's career and you can see the large increase since. There was also an increase post FOX and digital age. FOX has never broadcasted an NBA game and therefore had no impact on Sports Illustrated covers. The digital age increase can be chalked up to potentially the LeBron James era, which started in 2003/2004. However, there isn't enough data in terms of both years and LeBron's career still going to determine anything further about the numbers post digital age in that sense.

```{r}
model_filtered3 <- lm(`year` ~ `share`, data = basketball)
summary(model_filtered3)
```

This is the most significant result thus far. With an EXTREMELY low p-value and a high coefficient, the increase is both large and significant. This result shows me that while some sports trended down, others had to take their places. The NBA was one of those sports to take the spot of the declining NHL and boxing.





BASEBALL:

```{r, fig.width = 12, fig.height = 5}
vlines <- data.frame(
  year = c("1979", "1994", "2000"),
  event = c("Start of ESPN", "Start of FOX", "Start of Digital Age")
)

ggplot(baseball[baseball$year != "Grand Total", ], 
       aes(x = as.numeric(year), y = share, group = 1)) +
  geom_line() +
  geom_vline(data = vlines, aes(xintercept = as.numeric(year), color = event), linetype = "solid") +
  scale_color_manual(
    name = "Key Events",
    values = c("Start of ESPN" = "red", "Start of FOX" = "blue", "Start of Digital Age" = "green")
  ) +
  scale_x_continuous(breaks = seq(1955, 2013, by = 5)) +
  labs(
    x = "Year",
    y = "Number of Baseball Covers",
    title = "Sports Illustrated Covers: Baseball Coverage Over Time"
  ) +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1, size = 9)
  )
```
The MLB graph is very up and down. Based on the graph alone, there is a slight increase, but we will see further down. The peak around 1998 corresponds with the Yankees superteam consisting of Derek Jeter, Mariano Rivera, Bernie Williams, Paul O'Neill, and more. The other peaks correspond with big years for the sport such as 2004 when the Red Sox won their first World Series in 86 years and 2009 when the Yankees won their most recent and MLB leading 27th title. The graph is not consistent in any way. The up and down look makes it seem like there is no real change over time post 1980. 

```{r}
baseball_filtered <- baseball[baseball$year != "Grand Total", ]
baseball_filtered$year <- as.numeric(baseball_filtered$year)
baseball_filtered$count <- as.numeric(baseball_filtered$share)

# Before/After ESPN (1979)
baseball_filtered$espn <- ifelse(baseball_filtered$year < 1979, "Before ESPN", "After ESPN")
aggregate(count ~ espn, data = baseball_filtered, FUN = mean)

# Before/After FOX (1994)
baseball_filtered$fox <- ifelse(baseball_filtered$year < 1994, "Before FOX", "After FOX")
aggregate(count ~ fox, data = baseball_filtered, FUN = mean)

# Before/After Digital Age (2000)
baseball_filtered$digital <- ifelse(baseball_filtered$year < 2000, "Before Digital", "After Digital")
aggregate(count ~ digital, data = baseball_filtered, FUN = mean)
```
The increase in share post ESPN and FOX are due to the two networks broadcasting America's Pastime. While the graphs look up and down, however, the differences are so minimal that it just about averages out. There isn't much to take away from this test here.

```{r}
model_filtered4 <- lm(`year` ~ `share`, data = baseball)
summary(model_filtered4)
```
This is another significant increase, according to this regression model. While it's not very visible on the graph, the underlying data is there to show that this is a significant increase.




AMERICAN FOOTBALL:

```{r, fig.width = 12, fig.height = 5}
vlines <- data.frame(
  year = c("1979", "1994", "2000"),
  event = c("Start of ESPN", "Start of FOX", "Start of Digital Age")
)

ggplot(american_football[american_football$year != "Grand Total", ], 
       aes(x = as.numeric(year), y = share, group = 1)) +
  geom_line() +
  geom_vline(data = vlines, aes(xintercept = as.numeric(year), color = event), linetype = "solid") +
  scale_color_manual(
    name = "Key Events",
    values = c("Start of ESPN" = "red", "Start of FOX" = "blue", "Start of Digital Age" = "green")
  ) +
  scale_x_continuous(breaks = seq(1955, 2013, by = 5)) +
  labs(
    x = "Year",
    y = "Number of American Football Covers",
    title = "Sports Illustrated Covers: American Football Coverage Over Time"
  ) +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1, size = 9)
  )
```
Like the NBA, this increase is obvious on the graph. However, there are less valleys on the NFL graph. This one is a very steady increase over time. The peaks in the early 2000s correlate with the big boom in popularity for the NFL and fantasy football. Sports Illustrated picks their cover athletes based on what the biggest story at the time is. The NFL is usually the largest sport at any given time of the year and this is reflected in this data.

```{r}
american_football_filtered <- american_football[american_football$year != "Grand Total", ]
american_football_filtered$year <- as.numeric(american_football_filtered$year)
american_football_filtered$count <- as.numeric(american_football_filtered$share)

# Before/After ESPN (1979)
american_football_filtered$espn <- ifelse(american_football_filtered$year < 1979, "Before ESPN", "After ESPN")
aggregate(count ~ espn, data = american_football_filtered, FUN = mean)

# Before/After FOX (1994)
american_football_filtered$fox <- ifelse(american_football_filtered$year < 1994, "Before FOX", "After FOX")
aggregate(count ~ fox, data = american_football_filtered, FUN = mean)

# Before/After Digital Age (2000)
american_football_filtered$digital <- ifelse(american_football_filtered$year < 2000, "Before Digital", "After Digital")
aggregate(count ~ digital, data = american_football_filtered, FUN = mean)
```
The increases for post all of these lines in my opinion don't reflect the broadcast partners. Sure all of these networks broadcast the NFL, but the sport overall was increasing so quickly in popularity that these numbers are almost irrelevant aside from getting more eyes on the sport from large TV networks. 


```{r}
model_filtered5 <- lm(`year` ~ `share`, data = american_football)
summary(model_filtered5)
```
This is another significant increase in coverage. The NFL increased steadily over time in terms of both general popularity and Sports Illustrated cover shares.


Story Pitch:
What Sports Illustrated Covers Tell Us About America

For 59 years, Sports Illustrated made a decision every single week: whose face goes on the cover? That choice wasn't made in a vacuum. It reflected what editors believed Americans cared about most at that exact moment in time. Taken together, those thousands of decisions form an accidental but remarkably precise record of how American sports culture and American identity evolved.

The article would cover the NFL's dominance in US Mainstream Media, the decline of boxing, the NBA's steady growth, the MLB's up and down history and the NHL's lagging behind.

Potential Problems
Data ends in 2013, no streaming era, no social media, individual athletes skew the data further than sports.

Reporting Plan
Talk to Sports Illustrated editors, sports media historians. Use the charts I made in this project and create others as well.
