---
title: "final_project"
output: html_document
---
The dataframe contains data from 1954-2013. The data shows the different sports that were featured on the cover of Sports Illustrated over that 59 year span. The columns that are specific to my project are: baseball, basketball, boxing, american football and ice hockey. This project will look at their coverage on the magazine cover over time, being the "Big Four" sports in the US. How do they fluctuate in comparison to each other? How does this reflect the popularity of each sport? I will exclude the data from 1954 because they only printed 20 issues that year, rather than the 52 or 53 from every other year in the dataset.



Hypothesis testing to see the correlation over time of how commonly they were used. Hypotheses: Football became more common of a cover sport over time. Baseball and basketball remained relatively flat but on a slight decline. Ice hockey was kind of randomly thrown in there over time when deemed right. To help with this, I will create bar charts to look at whether or not these sports have a normal distribution over time, if they are skewed towards the earlier or later years, or if it is random. This would be used to help with the hypothesis testing.

Fox started broadcasting sports in 1994. How did the broadcasting company bringing more eyes to live sports affect who was on the cover of the magazine? I also want to do the same thing, but for ESPN. They started in 1979. To avoid having one side of the comparison have more data than the other greatly affecting the numbers, I will average out the results in addition to having the raw numbers to get a more accurate representation.

I want to do a similar comparison, but using the year 2000. I chose this year as well because I want to see if the "digital age" had any affect on which sports were featured more frequently.

Linear regression would be the perfect way to test the above comparisons (FOX, ESPN, 2000). I will do a similar comparison to what we did in class with the FARMS data to see if there is a correlation over time.

Outlier analysis. Were there any years where there was little to no coverage of any of these sports? Or on the flip side, were there any years where there was a much larger number of issues made with the sport featured? What could have caused these outliers?

I also want to turn each data entry into a percentage of the total issues for that year. How do they compare to each other?

```{r}

library(tidyverse)
library(janitor)
library(dplyr)


covers <- read_csv("SI-Cover-by-Sport-DFE.csv")

covers <- covers %>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))

```

```{r}
covers <- read_csv("SI-Cover-by-Sport-DFE.csv") |> clean_names()  |> rename (year = unnamed_0) # becomes: american_football, ice_hockey, grand_total, etc.

covers2 <- covers |>
  filter(year >= 1955)
covers2 <- covers %>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))

```

```{r}
covers2 <- covers2 |>
   mutate(across(c(american_football, baseball, boxing, basketball, ice_hockey),
                ~replace_na(., 0))) |>
  pivot_longer(c(american_football, baseball, boxing, basketball, ice_hockey),
               names_to = "sport", values_to = "count") |>
  mutate(share = count / grand_total)
  
```

```{r}
covers2 <- covers2 |>
  filter(year >= 1955)
```

```{r}
covers2 <- covers2 |>
  select(year, grand_total, sport, count, share)
```

I have modified the dataframe to now only show the data that is specific for my project. I then filtered it by "share" to see what the highest shares of covers by sport were from 1955-2013. The highest percentage is 46% for American Football in 2005. In fact, each of the top 12 cover share percentages are all American Football, showing it's prominence and dominance in the sports landscape. The next 12 are all tied at 17 covers in the year. 8 of them are American Football. This means that of the top 24, 16.66% of them were not American Football, meaning 83.33% of them were. This was 4 of the 24. 3 of them were basketball and 1 was baseball.

It is the same story when looking at the overall numbers. 27% of all Sports Illustrated covers from 1955-2013 had American Football represented on the cover. 19% baseball. 18% basketball. 4% boxing. 3% ice hockey. 71% of all Sports Illustrated covers were one of these 5 sports, which are arguably the top 5 sports in US history. Other sports like NASCAR, golf and soccer have their popularity too, but when considering all of US history, the 5 sports I mentioned above are at the top of the charts.


